---
title: "Biodiversity_Indicators"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Biodiversity_Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

# Libraries
library(tidyverse)
library(Rmisc)
library(ggthemes)
library(readxl)
library(msk)

# Plot settings
theme_set(
  theme_clean() +
    theme(
      legend.title = element_blank(), 
      legend.position = "right", 
      legend.background = element_rect(colour = "white"),
      plot.background = element_blank())
)
options(ggplot2.discrete.colour= c("#1F78B4", "#FF7F00", "#33A02C", "#E31A1C", "#6A3D9A"))
```

# Data preparation
The data structure adheres to the [Darwin Core](https://dwc.tdwg.org/) standart, which is utilised by numerous biodiversity data providers, such as GBIF.

A dataset consists of at least the following two tables (classes according to Darwin Core):

[Event](https://dwc.tdwg.org/terms/#event): Table for all surveys 

Contains one row for each survey. This table is the main table (core class according to Darwin Core) and must include a column `eventID`  with unique IDs. All other columns are optional and can be added as needed:

- `locationID` : If the dataset includes a table for the survey locations, a locationID is essential in order to associate the events with the survey locations.
- Additional columns, such as the date of collection, can be added as needed.

[Occurence](https://dwc.tdwg.org/terms/#occurrence): Table for the recorded taxa

The documentation of an organism at a specific event serves as the basis for calculating the indicators. The `event ID`  is essential for associating the recorded taxa with individual surveys. Z

Zusätzlich sind zwei weitere Tabellen möglich

[Location](https://dwc.tdwg.org/terms/#location): Table for the data collection location.

Enthält für jede Erhebung eine Zeile. Diese Tabelle Ist die Haupttabelle (core class gemäss darwin core) und muss eine Spalte "mit einer eindeutige `locationID`  für die erhebung enthalten. Weitere Spalten wie das Datum der Erhebung können beliebig ergänzt werden. Für jede 

locationID

Taxon: Species Lists

# Calculate indicators
For the calculation, there is a separate function for each indicator. These can be chained sequentially.

```{r}
dat <- 
  dia %>% 
  calc_alphadiv()
```

Die berechneten Indikatorwerte sind nun in `events` gespeichert. 

```{r}
dat$event
```

# Resultate plotten

Jährliche Mittelwerte berechen und plotten

```{r}
dat$event %>% 
  group_by(year = year(eventDate)) %>% 
  dplyr::summarise(
    av = mean(alphadiv)
  ) %>% 
  ggplot(aes(x = year, y = av)) +
  geom_point() +
  geom_smooth(method = lm) +
  ylim(0, NA) +
  labs(
    x = "Species richness",
    y = ""
  )
```




