---
title: "Calculating Biodiversity Indicators with msk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Biodiversity Indicators with msk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

The `msk` package provides a collection of R functions for calculating
biodiversity indicators from species occurrence data. It was developed for
analysing data from the Swiss
[Modular Stepwise Procedure](https://modul-stufen-konzept.ch/en/about-the-msp/)
(Modulstufenkonzept, MSK), a set of standardised methods for surveying and
assessing the status of water bodies in Switzerland. However, the functions
work with any dataset that follows the
[Darwin Core](https://dwc.tdwg.org/terms/) standard.

The package currently provides functions for:

- **Alpha diversity** -- species richness per sampling event
- **Total abundance** -- sum of individuals per sampling event
- **Gamma diversity** -- total species richness across events
- **Beta diversity** -- Simpson dissimilarity between events
- **Index normalisation** -- express indicator values relative to a reference

```{r setup, message = FALSE}
library(msk)
library(dplyr)
library(ggplot2)
```

# Data structure

The `msk` package expects data as a **list** with two data frames (tibbles),
following the [Darwin Core](https://dwc.tdwg.org/terms/) standard:

**`event`** -- one row per sampling event (Darwin Core *Event* class):

| Column       | Description                          | Required |
|:-------------|:-------------------------------------|:---------|
| `eventID`    | Unique identifier for the event      | yes      |
| `eventDate`  | Date of the sampling event           | no       |
| `locationID` | Identifier for the sampling location | no       |

**`occurrence`** -- one row per taxon observation (Darwin Core *Occurrence* class):

| Column            | Description                            | Required |
|:------------------|:---------------------------------------|:---------|
| `eventID`         | Link to the event table                | yes      |
| `taxonID`         | Identifier for the recorded taxon      | yes      |
| `individualCount` | Number of individuals observed         | no       |

The two tables are linked by `eventID`. Additional columns following Darwin
Core terminology can be added freely (e.g. `samplingProtocol`, `habitat`,
`scientificName`).

## Example dataset: Diatoms

The package includes a diatom dataset (`dia`) from the Swiss NAWA monitoring
programme:

```{r}
data("dia")
str(dia, max.level = 1)
```

The **event** table contains `r nrow(dia$event)` sampling events:

```{r}
head(dia$event)
```

The **occurrence** table contains `r nrow(dia$occurrence)` taxon records:

```{r}
head(dia$occurrence)
```

# Calculating indicators

All indicator functions follow the same pattern: they take a data list `dat`
and return it with the computed indicator added. This allows chaining
multiple calculations.

## Alpha diversity

Alpha diversity is the number of distinct taxa recorded at a single sampling
event. Use `calc_alphadiv()` to compute it:

```{r}
dat <- dia %>%
  calc_alphadiv()

head(dat$event)
```

The new column `alphadiv` contains the species richness per event.

## Total abundance

Total abundance is the sum of `individualCount` per event:

```{r}
dat <- dat %>%
  calc_abundance()

head(dat$event)
```

Both indicators can be computed in a single chain:

```{r, eval = FALSE}
dat <- dia %>%
  calc_alphadiv() %>%
  calc_abundance()
```

## Gamma diversity

Gamma diversity is the total number of distinct taxa across all events
(or within groups). Unlike alpha diversity, this is a group-level metric,
so `calc_gammadiv()` returns a separate summary table rather than adding a
column to the event table.

```{r}
# Overall gamma diversity
calc_gammadiv(dia)
```

To compute gamma diversity per year, first add a year column to the event
table, then pass the grouping column name:

```{r}
dat$event$year <- as.integer(format(dat$event$eventDate, "%Y"))

gamma_by_year <- calc_gammadiv(dat, group_by_col = "year")
gamma_by_year
```

## Beta diversity

Beta diversity measures how different communities are between sampling events.
The `calc_betadiv()` function computes pairwise Simpson dissimilarity:

$$\beta_{sim} = \frac{\min(b, c)}{a + \min(b, c)}$$

where *a* is the number of shared taxa, *b* and *c* are the numbers of taxa
unique to each event. The index ranges from 0 (identical) to 1 (completely
different) and is independent of richness differences.

```{r}
# Overall beta diversity
calc_betadiv(dia)
```

```{r}
# Beta diversity per year
beta_by_year <- calc_betadiv(dat, group_by_col = "year")
beta_by_year
```

# Visualising results

The indicator outputs are standard tibbles that work seamlessly with
`ggplot2`.

## Alpha diversity over time

```{r alpha-plot}
alpha_summary <- dat$event %>%
  group_by(year) %>%
  summarise(
    mean = mean(alphadiv),
    lo = mean - qt(0.975, n() - 1) * sd(alphadiv) / sqrt(n()),
    up = mean + qt(0.975, n() - 1) * sd(alphadiv) / sqrt(n())
  )

ggplot(alpha_summary, aes(x = year, y = mean, ymin = lo, ymax = up)) +
  geom_point() +
  geom_errorbar(width = 0.2) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  ylim(0, NA) +
  labs(
    title = "Alpha Diversity",
    x = "",
    y = "Number of taxa per event"
  )
```

## Gamma diversity over time

```{r gamma-plot}
ggplot(gamma_by_year, aes(x = year, y = gammadiv)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  ylim(0, NA) +
  labs(
    title = "Gamma Diversity",
    x = "",
    y = "Total number of taxa"
  )
```

## Beta diversity over time

```{r beta-plot}
ggplot(beta_by_year, aes(x = year, y = betadiv_mean,
                         ymin = betadiv_q25, ymax = betadiv_q75)) +
  geom_point() +
  geom_errorbar(width = 0.2) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  ylim(0, NA) +
  labs(
    title = "Beta Diversity (Simpson Dissimilarity)",
    x = "",
    y = "Community dissimilarity"
  )
```

# Index normalisation

To compare indicators on a common scale, use `calc_indicator_index()` to
normalise values relative to a reference year (base = 100):

```{r}
gamma_indexed <- calc_indicator_index(
  gamma_by_year,
  value_col = "gammadiv",
  group_col = "year",
  ref_value = min(gamma_by_year$year)
)
gamma_indexed
```

```{r index-plot}
ggplot(gamma_indexed, aes(x = year, y = index)) +
  geom_hline(yintercept = 100, linetype = 2) +
  geom_point() +
  geom_line() +
  labs(
    title = "Gamma Diversity Index",
    x = "",
    y = paste0("Index [", min(gamma_by_year$year), " = 100]")
  )
```

# Extending the package

The `msk` package is designed to be extended with additional indicator
functions. Planned additions include:

- **IBCH** (Indice Biologique Global Harmonisé) -- for macroinvertebrate
  assessment
- **SPEAR Index** -- pesticide sensitivity indicator
- **Shannon diversity / Evenness** -- abundance-weighted diversity measures
- **Trend analysis** -- temporal trend estimation with mixed models

New indicator functions should follow the established pattern:

1. Take a `dat` list with `event` and `occurrence` tables
2. Compute the indicator
3. Return the modified `dat` list (for event-level indicators) or a summary
   tibble (for group-level indicators)

This ensures all functions are composable and work with the same data
structure.
