---
title: "Calculating Biodiversity Indicators with msk"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Calculating Biodiversity Indicators with msk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

The `msk` package provides R functions for calculating biodiversity indicators
from species occurrence data. It was developed for analysing data from the Swiss
[Modular Stepwise Procedure](https://modul-stufen-konzept.ch/en/about-the-msp/)
(Modulstufenkonzept, MSK), but works with any dataset following the
[Darwin Core](https://dwc.tdwg.org/terms/) standard.

This vignette is structured in three parts:

1. **Data structure** — how to organise your data for use with `msk`
2. **Indicator functions** — how each function works, illustrated with
   macroinvertebrate data (`mzb`)
3. **Complete example** — a full analysis of Swiss NAWA fish monitoring data
   (`fisch`), including a custom biomass metric

```{r setup, message = FALSE}
library(msk)
library(dplyr)
library(ggplot2)
library(patchwork)
```

---

# Data structure

The `msk` package expects data as a **list** with two data frames (tibbles),
following the [Darwin Core](https://dwc.tdwg.org/terms/) standard.

**`event`** — one row per sampling event:

| Column       | Description                          | Required |
|:-------------|:-------------------------------------|:---------|
| `eventID`    | Unique identifier for the event      | yes      |
| `eventDate`  | Date of the sampling event           | no       |
| `locationID` | Identifier for the sampling location | no       |

**`occurrence`** — one row per taxon observation:

| Column            | Description                            | Required |
|:------------------|:---------------------------------------|:---------|
| `eventID`         | Link to the event table                | yes      |
| `taxonID`         | Identifier for the recorded taxon      | yes      |
| `individualCount` | Number of individuals observed         | no       |

The two tables are linked by `eventID`. Additional columns (e.g. `year`,
`altitude`, `scientificName`) can be added freely — they are preserved by all
package functions.

## The `mzb` dataset

The package includes a macroinvertebrate dataset (`mzb`) from the Swiss NAWA
monitoring programme, covering four campaigns (2012, 2015, 2019, 2023) across
85 stations below 1000 m a.s.l. that were sampled in all four campaigns:

```{r}
data("mzb")
str(mzb, max.level = 1)
```

The **event** table has one row per sampling event:

```{r}
head(mzb$event)
```

The **occurrence** table has one row per taxon per event:

```{r}
head(mzb$occurrence)
```

---

# Indicator functions

The `msk` package provides two types of indicator functions:

- **Event-level functions** add a new column to the `event` table and return
  the modified data list. They are chainable with `%>%`.
- **Group-level functions** compute indicators across sets of events and return
  a separate summary tibble.

## Alpha diversity

Alpha diversity is the number of distinct taxa per sampling event.
`calc_alphadiv()` adds an `alphadiv` column to the event table:

```{r}
dat <- mzb %>%
  calc_alphadiv()

head(dat$event)
```

## Abundance

`calc_abundance()` supports two methods, selected with the `method` argument:

- **`"total"`** (default): sum of `individualCount` per event. Reflects overall
  abundance but is sensitive to dominant taxa.
- **`"mean_per_taxon"`**: mean of per-taxon counts per event. Less sensitive
  to dominant taxa; useful for tracking whether individual species become more
  or less abundant on average across taxa of different commonness.

Both methods can be applied in sequence:

```{r}
dat <- dat %>%
  calc_abundance(method = "total") %>%
  calc_abundance(method = "mean_per_taxon")

head(dat$event)
```

The result columns are named `abundance` (total) and `abundance_mean`
(mean per taxon).

## Gamma diversity

Gamma diversity is the total number of distinct taxa across a group of events.
Unlike alpha diversity, this is a group-level metric, so `calc_gammadiv()`
returns a summary tibble. Use `group_by_col` to compute it separately per year:

```{r}
calc_gammadiv(mzb, group_by_col = "year")
```

## Beta diversity

Beta diversity measures how different communities are between events.
`calc_betadiv()` computes pairwise **Simpson dissimilarity**:

$$\beta_{sim} = \frac{\min(b, c)}{a + \min(b, c)}$$

where *a* = shared taxa, *b* and *c* = taxa unique to each event. The index
ranges from 0 (identical communities) to 1 (no shared taxa) and is independent
of richness differences.

```{r}
calc_betadiv(mzb, group_by_col = "year")
```

The output contains the mean and the 25th/75th percentile of pairwise
dissimilarities within each group.

---

# Chaining and visualisation

Event-level functions return the data list unchanged (except for the new
column), so they compose naturally. The following pipeline computes alpha
diversity and abundance for `mzb` in a single chain:

```{r mzb-calc}
campaigns <- c(2012, 2015, 2019, 2023)

mzb_ind <- mzb %>%
  calc_alphadiv() %>%
  calc_abundance(method = "total")

alpha_mzb <- mzb_ind$event %>%
  group_by(year) %>%
  summarise(
    mean = mean(alphadiv),
    lo   = mean - qt(0.975, n() - 1) * sd(alphadiv) / sqrt(n()),
    up   = mean + qt(0.975, n() - 1) * sd(alphadiv) / sqrt(n()),
    .groups = "drop"
  )

abund_mzb <- mzb_ind$event %>%
  group_by(year) %>%
  summarise(
    mean = mean(abundance),
    lo   = mean - qt(0.975, n() - 1) * sd(abundance) / sqrt(n()),
    up   = mean + qt(0.975, n() - 1) * sd(abundance) / sqrt(n()),
    .groups = "drop"
  )

beta_mzb <- calc_betadiv(mzb, group_by_col = "year")
```

All indicator outputs are standard tibbles and work directly with `ggplot2`:

```{r mzb-plot, fig.width = 10, fig.height = 4}
p_alpha <- ggplot(alpha_mzb, aes(x = year, y = mean, ymin = lo, ymax = up)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Artenvielfalt\n(Alpha-Diversität)",
    x     = "Zeitperiode",
    y     = "Artenvielfalt\n[Anzahl Taxa pro Probestelle]"
  )

p_beta <- ggplot(beta_mzb, aes(x = year, y = betadiv_mean,
                                ymin = betadiv_q25, ymax = betadiv_q75)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Vielfalt der Artgemeinschaften\n(Beta-Diversität)",
    x     = "Zeitperiode",
    y     = "Vielfalt der Artgemeinschaften\n[Simpson Index (0–1)]"
  )

p_abund <- ggplot(abund_mzb, aes(x = year, y = mean, ymin = lo, ymax = up)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Abundanz",
    x     = "Zeitperiode",
    y     = "Abundanz\n[Gesamtindividuenzahl pro Probestelle]"
  )

p_alpha + p_beta + p_abund
```

---

# Index normalisation

To express indicators relative to a reference point (base = 100), use
`calc_indicator_index()`. This is useful for comparing trends across indicators
or datasets measured on different scales.

Here we normalise gamma diversity relative to the 2012 baseline:

```{r index-calc}
gamma_mzb <- calc_gammadiv(mzb, group_by_col = "year")

gamma_indexed <- calc_indicator_index(
  gamma_mzb,
  value_col = "gammadiv",
  group_col = "year",
  ref_value = 2012
)
gamma_indexed
```

```{r index-plot}
ggplot(gamma_indexed, aes(x = year, y = index)) +
  geom_hline(yintercept = 100, linetype = 2) +
  geom_point(size = 2) +
  geom_line() +
  scale_x_continuous(breaks = campaigns) +
  labs(
    title = "Gamma-Diversität MZB (Index)",
    x     = "Zeitperiode",
    y     = "Index [2012 = 100]"
  )
```

---

# Complete example: Fish data

The `fisch` dataset contains NAWA fish monitoring data from four campaigns
(2012, 2015, 2019, 2023) across 51 stations sampled in at least three of the
four campaigns. It demonstrates how the Darwin Core structure accommodates
additional variables: the `totalWeight_g` column in the occurrence table
enables biomass analyses alongside the standard diversity indicators.

## Data overview

```{r}
data("fisch")
str(fisch, max.level = 1)
```

The event table contains metadata including the number of electrofishing runs
(`N_runs`) and the stream name:

```{r}
head(fisch$event)
```

The occurrence table contains one row per species per event, with abundance
(individuals), biomass, and species names:

```{r}
head(fisch$occurrence)
```

## Calculating indicators

We use the standard `msk` functions for alpha diversity and abundance, then
add a custom biomass metric derived directly from the occurrence table. This
illustrates how the Darwin Core structure makes it straightforward to extend
the standard workflow with dataset-specific metrics:

```{r fisch-calc}
fisch_ind <- fisch %>%
  calc_alphadiv() %>%
  calc_abundance(method = "total")

# Custom metric: total biomass per event (sum of totalWeight_g)
biomass_per_event <- fisch$occurrence %>%
  group_by(eventID) %>%
  summarise(biomass_g = sum(totalWeight_g, na.rm = TRUE), .groups = "drop")

fisch_ind$event <- fisch_ind$event %>%
  left_join(biomass_per_event, by = "eventID")

head(fisch_ind$event)
```

## Summarising by year

```{r fisch-summary}
campaigns <- c(2012, 2015, 2019, 2023)

alpha_fisch <- fisch_ind$event %>%
  group_by(year) %>%
  summarise(
    mean = mean(alphadiv),
    lo   = mean - qt(0.975, n() - 1) * sd(alphadiv) / sqrt(n()),
    up   = mean + qt(0.975, n() - 1) * sd(alphadiv) / sqrt(n()),
    .groups = "drop"
  )

abund_fisch <- fisch_ind$event %>%
  group_by(year) %>%
  summarise(
    mean = mean(abundance),
    lo   = mean - qt(0.975, n() - 1) * sd(abundance) / sqrt(n()),
    up   = mean + qt(0.975, n() - 1) * sd(abundance) / sqrt(n()),
    .groups = "drop"
  )

biomass_fisch <- fisch_ind$event %>%
  group_by(year) %>%
  summarise(
    mean = mean(biomass_g),
    lo   = mean - qt(0.975, n() - 1) * sd(biomass_g) / sqrt(n()),
    up   = mean + qt(0.975, n() - 1) * sd(biomass_g) / sqrt(n()),
    .groups = "drop"
  )

beta_fisch  <- calc_betadiv(fisch, group_by_col = "year")
gamma_fisch <- calc_gammadiv(fisch, group_by_col = "year")
```

## Results

```{r fisch-plot, fig.width = 10, fig.height = 8}
p_alpha_f <- ggplot(alpha_fisch, aes(x = year, y = mean, ymin = lo, ymax = up)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Artenvielfalt\n(Alpha-Diversität)",
    x     = "Zeitperiode",
    y     = "Artenvielfalt\n[Anzahl Arten pro Probestelle]"
  )

p_abund_f <- ggplot(abund_fisch, aes(x = year, y = mean, ymin = lo, ymax = up)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Abundanz",
    x     = "Zeitperiode",
    y     = "Abundanz\n[Gesamtindividuenzahl pro Probestelle]"
  )

p_biomass_f <- ggplot(biomass_fisch, aes(x = year, y = mean, ymin = lo, ymax = up)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Biomasse",
    x     = "Zeitperiode",
    y     = "Biomasse\n[Gesamtgewicht g pro Probestelle]"
  )

p_beta_f <- ggplot(beta_fisch, aes(x = year, y = betadiv_mean,
                                    ymin = betadiv_q25, ymax = betadiv_q75)) +
  geom_point() +
  geom_errorbar(width = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Vielfalt der Artgemeinschaften\n(Beta-Diversität)",
    x     = "Zeitperiode",
    y     = "Vielfalt der Artgemeinschaften\n[Simpson Index (0–1)]"
  )

(p_alpha_f + p_abund_f) / (p_biomass_f + p_beta_f)
```

Gamma diversity shows the total number of fish species recorded across all
stations per campaign:

```{r fisch-gamma, fig.width = 5, fig.height = 4}
ggplot(gamma_fisch, aes(x = year, y = gammadiv)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2) +
  scale_x_continuous(breaks = campaigns) +
  ylim(0, NA) +
  labs(
    title = "Gamma-Diversität Fische",
    x     = "Zeitperiode",
    y     = "Gesamtartenzahl"
  )
```

